---
title: "lesson-06: The Language of Probability"
output: html_notebook
---

```{r}
library(tidyverse)
library(magrittr)
```

# The Language of Probability

## Outcomes

Probability models can be applied to any situation in which there are multiple potential outcomes and there is uncertainty about which outcome will occur. Due to the wide variety of types of random phenomena, an **outcome** can be virtually anything:

-   the result of a coin flip
-   the results of a sequence of coin flips
-   a shuffle of a deck of cards
-   the weather conditions tomorrow in your city
-   the path of a particular Atlantic hurricane

And so on.

Before the random phenomenon occurs it is unknown which outcome will be the result. When the phenomenon takes place, a particular outcome is observed. The first step in defining a probability model for a random phenomenon is to identify the possible outcomes. The **sample space** is the collection of all possible outcomes of a random phenomenon.

### Example

Roll a four-sided die twice, and record the result of each roll in sequence. For example, a 3 on the first roll and a 1 on the second is not the same outcome as a 1 on the first roll and a 3 on the second.

Identify the sample space.

We might be interested in the sum of the two rolls. Explain why it is still advantageous to define the sample space as in the previous part, rather than as just 2, ..., 8.

```{r}
tibble(
  x = seq(1, 4),
  y = seq(1, 4)
) %>%
  expand(., x, y)
```

### Example

Consider the matching problem with `n = 4`. Label the objects 1, 2, 3, 4, and the spots 1, 2, 3, 4, and assume that any objects can go into any of the available spots. Identify an appropriate sample space.

```{r}
tibble(
  spot_1 = seq(1, 4),
  spot_2 = seq(1, 4),
  spot_3 = seq(1, 4),
  spot_4 = seq(1, 4)
) %>%
  expand(., spot_1, spot_2, spot_3, spot_4) %>%
  mutate(., i = row_number()) %>%
  group_by(., i) %>%
  mutate(., s = anyDuplicated(c(spot_1, spot_2, spot_3, spot_4))) %>%
  ungroup() %>%
  filter(., s == 0)
```

or, using `rowwise()`,

```{r}
tibble(
  spot_1 = seq(1, 4),
  spot_2 = seq(1, 4),
  spot_3 = seq(1, 4),
  spot_4 = seq(1, 4)
) %>%
  expand(., spot_1, spot_2, spot_3, spot_4) %>%
  rowwise() %>%
  mutate(., s = anyDuplicated(c(spot_1, spot_2, spot_3, spot_4))) %>%
  filter(., s == 0)
```

## Events

An event is something that might happen. For example, if we’re interested in the weather conditions in our city tomorrow, events include

-   the high temperature is 75°F
-   the high temperature is above 75°F
-   it rains
-   it does not rain
-   it rains and the high temperature is above 75°F

An outcome consists of all the information about tomorrow’s weather conditions, while an event is a collection of outcomes that satisfy some criteria.

The sample space is the collection of all possible outcomes; an event represents only those outcomes which satisfy some criteria. Events are typically denoted with capital letters near the start of the alphabet, with or without subscripts (e.g. $A$, $B$, $A_1$, $B_2$, and so on).

If the outcomes of a sample space are represented by rows in a table, then events are collections (or subsets) or rows which satisfy some criteria.

-   Complements: $A^c$, or "not $A$"; outcomes that do not satisfy $A$.

-   Intersections: $A \cap B$, or "$A$ and $B$"; outcomes that satisfy both $A$ and $B$.

-   Unions: $A \cup B$, or "$A$ or $B$"; outcomes that satisfy $A$ and/or $B$.

### Example

Roll a four-sided die twice, and record the result of each roll in sequence. Using the sample space from before, identify the following events.

-   $A$, or the event where the sum of the two dice is 4.
-   $B$, or the event that the sum of the two dice is at most 3.
-   $C$, or the event that the larger of the two rolls (or the common roll if a tie) is 3.
-   $A \cup C$ (what does this mean?)
-   $D$, or the event that the first roll is a 3.
-   $E$, or the event that the second roll is a 3.
-   $D \cup E$ (what does this mean?)
-   $D \cap E$ (what does this mean?)
-   If the outcome is (1, 3), which of the events above occurred?

## Random Variables

We have already talked about random variables and uncertainties, so I will skip the explanation.

### Example

Roll a four-sided die twice, and record the result of each roll in sequence. Recall the sample space from above. Let $X$ be the sum of the two dice, and $Y$ be the larger of the two rolls (or the common value if both rolls are the same).

-   Construct a table identifying the values of $X$ and $Y$ for each outcome in the sample space.
-   Identify the possible values of $X$.
-   Identify the possible values of $Y$
-   Identify the possible values of the pair ($X$, $Y$)

```{r}
tibble(
  r1 = seq(1, 4),
  r2 = seq(1, 4)
) %>%
  expand(., r1, r2) %>%
  rowwise(.) %>%
  mutate(., x = sum(r1, r2), y = max(r1, r2))
```

\`\`\`
