---
title: "lesson-05: Learning Statistics with R"
output: html_notebook
---

# Basic Statistics with R

```{r}
library(tidyverse)
library(magrittr)
```


## Frequentist Approach

As an exercise, let's perform a frequentist simulation. Suppose you have a 6-sided die. What is the probability of having a number that is larger than, or equal to, 5? Take a large number of samples (let's say 1e4 samples) to verify your answer, and plot the resulting probability as a function of the number of samples.

```{r}
toss <- function(size = 10) {
  x <- sample(6, size = size, replace = TRUE) %>%
    {sum(. > 3) / length(.)}
}

pulls <- seq(1, 1e4, 50)
res <- sapply(pulls, toss)

df <- tibble(pulls = pulls, res = res)

print(tail(res, n = 1))

df %>%
  ggplot(., aes(x = pulls, y = res)) +
    geom_point() +
    geom_line() +
    geom_smooth()
```

## Probability

Look at the *garden of forking data*.

### Exercise

Suppose the following observations (samples). Define the possible proportions of `W`, and all the possible ways the observation could be made, then calculate the probabilities for each possibility.

```{r}
samples <- c("W", "L", "W", "W", "W", "L", "W", "L", "W")
```

Here is the solution.

```{r}
w <- sum(samples == "W") # Number of W observed
l <- sum(samples == "L") # Number of L observed

p <- c(0, 0.25, 0.5, 0.75, 1) # All possible realizations of W

ways <- function(q) {
  (q * 4)^w * ((1 - q) * 4)^l
}

counts <- sapply(p, ways)
prob <- counts / sum(counts)

df <- tibble(p, counts, prob)
df
```

And the corresponding histogram.

```{r}
ggplot(df, aes(p, prob)) +
  geom_col(width = 0.1)
```

```{r}
sim_globe <- function(p = 0.7, N = 9) {
  sample(c("W", "L"), size = N, prob = (c(p, 1 - p)), replace = TRUE)
}
```

```{r}
replicate(sim_globe(p = 0.5, N = 9), n = 10)
```

```{r}
compute_posterior <- function(samples, poss = c(0, 0.25, 0.5, 0.75, 1)) {
  w <- sum(samples == "W") # Number of W observed
  l <- sum(samples == "L") # Number of L observed

  ways <- function(q) {
    (q * 4)^w * ((1 - q) * 4)^l
  }

  counts <- sapply(poss, ways)
  post <- counts / sum(counts)
  bars <- sapply(poss, ways)

  tibble(poss, counts, post = round(post, 3))
}
```

```{r}
compute_posterior(sim_globe())
```

```{r}
tosses <- sim_globe()
print(tosses)

d <- tibble(n_trials  = 1:9, n_success = cumsum(tosses == "W"))

sequence_length <- 50

d %>% 
  expand_grid(p_water = seq(from = 0, to = 1, length.out = sequence_length)) %>%
  group_by(p_water) %>%
  mutate(
    lagged_n_trials = lag(n_trials, 1),
    lagged_n_success = lag(n_success, 1)) %>% 
  ungroup() %>% 
  mutate(
    prior = ifelse(
      n_trials == 1,
      .5,
      dbinom(
        x = lagged_n_success, 
        size = lagged_n_trials, 
        prob = p_water
      )
    ),
    likelihood = dbinom(
      x = n_success, 
      size = n_trials, 
      prob = p_water
    ),
    strip = str_c("n = ", n_trials)) %>%
  group_by(n_trials) %>% 
  mutate(prior      = prior / sum(prior),
         likelihood = likelihood / sum(likelihood)) %>%   
  
  # Plot
  ggplot(aes(x = p_water)) +
    geom_line(aes(y = prior), linetype = 2) +
    geom_line(aes(y = likelihood)) +
    scale_x_continuous("proportion water", breaks = c(0, .5, 1)) +
    scale_y_continuous("plausibility", breaks = NULL) +
    facet_wrap(~ strip, scales = "free_y")
```
