---
title: "Lesson 1: Basics of R Workflow"
output: html_notebook
---

# Getting Started with RStudio

## Install RStudio

Follow instructions on [this website](https://posit.co/download/rstudio-desktop/). The RStuido interface should be straightforward, but feel free to change the layout.

## Use R console to perform basic calculations

Using the R console menu, you can use R as a calculator. For example, try the following computations:

```{r}
1 + 1
```

The `[1]` in the output pane represents the index of the following value, which is `2` in this case. The index is useful when outputting a long list or a vector. For example,

```{r}
1:50
```

shows that the number in the square brackets corresponds to the 1-based index of the following output. You can always use this index to refer to the element in a vector, like the example below.

```{r}
a <- c(1:50)
a[[31]]
```

## Install Packages

Let's try to install packages through the R console. For this notebook, we will use the `lobstr` package. Run the following command to install this package.

```{r eval=FALSE}
install.packages("lobstr")
```

And load the package to this notebook.

```{r}
library(lobstr)
```

## Use Help Pages

It is a good idea to refer to documentation. Let's take a look at `c()`, for example.

```{r}
help(c)
```

## Create R Notebook Page

Try executing this chunk by clicking the *Run* button within the chunk or by placing your cursor inside it and pressing *Cmd+Shift+Enter*.

```{r}
plot(cars)
```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

# The Very Basics

## Assignments

When you create objects in R programming language, you make an **assignment** statement:

```{r eval=FALSE}
object_name <- value
```

which roughly translates to "assign some `value` to an object and name it as `object_name`" in plain English. However, a more precise definition is needed to avoid confusion.

Let's consider the following assignment.

```{r}
x <- c(1, 2, 3)
```

The above statement is in fact performing two things:

- Create an object consisting of a vector `[1, 2, 3]`.
- Bind the object above to a name `x`.

We will talk more about vectors later.

The *name* is a reference to an object. In R, the variable/object name is simply a reference, so you can define a reference to a name, which is a reference to the vector we created.

```{r}
y <- x
```

```{r}
obj_addr(x)
obj_addr(y)
```

Now, what happens when you change the value of `y`?

### Exercise

- Explain the relationship between `a`, `b`, `c` and `d` in the following code:

```{r}
a <- 1:10
b <- a
c <- b
d <- 1:10
```

- Consider the following code. It binds `x` and `y` to a single object, and it attempts to modify `y`

```{r}
x <- c(1, 2, 3)
y <- x

y[[3]] <- 4

x
y
```

Why did the above code not modify `x`?

## Vectors

There are two types of vectors: atomic vectors (sometimes simply called *vectors*) and lists (also called *generic vectors*). They differ in terms of the type(s) of elements they hold. Atomic vectors hold elements of the same type, and lists can hold elemtns of different types.

In this lesson, we will focus on (atomic) vectors.

### Atomic Vectors

There are four primary types of atomic vectors:

- Logical
- Integer
- Double
- Character (which contains strings)

And there are two rare types: complex and raw. 

### Vector Operatios

Vector operations in R always performs element-wise operations. R will not do linear algebra unless you explicitly ask it to.

Let's define a simple (atomic) vector.

```{r}
x <- c(1:6)
x
```

All operations are done element-wise, even the multiplication.

```{r}
x * x
```

If you give R two vectors of unequal lengths, R will repeat the shorter vector until it is as long as the longer vector -- which is a terrible thing to do.

```{r}
x * 1:2
```

At the very least, you will see a warning (although it should really be an error) if the length of the shorter vector is not an integer multiple of the longer vector; for example,

```{r}
x * 1:4
```

which is still ghastly.

It is still possible to perform traditional linear algebra operations, fortunately; inner products can be calculated with the `%*%` operator, and outer products with the `%o%` operator, as shown below.

```{r}
x %*% x
```

```{r}
x %o% x
```

### Exercise

Try to figure out how to transpose a matrix and calculate the determinant of that matrix; use the help function to locate the appropriate tools.